---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---
STEPS ALONG TWO MONTHS
======================
**_by Max Testa_**


```{r, echo=TRUE, results='markup'}
sessionInfo()
```


````{r, results='hide', echo=FALSE}
## sets the language to have week-days names in English
Sys.setlocale("LC_TIME", "en_US.UTF-8")
```

## Loading and preprocessing the data
The data, representing the observations of the number of steps in 5' intervals
recorded from 2012-10-01 to 2012-11-30, are stored in `activity.csv` file,
downloaded from [Project site: repdata_data_activity.zip](https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip) [52K]
and unzipped in the working directory.
Data are read into a dataframe called `activity`, with its own structure:
```{r loading_data, results='markup', echo=1:5} 
activity<-read.csv("activity.csv", header=TRUE, stringsAsFactor=FALSE)
str(activity)
```

Variable `interval` is transformed from an `integer` into a `4 digit string`
to make it more uniform and meaningful.
Variable `date` is coerced to class "Date".
```{r setting_variable_format_&_class, results='markup', echo=TRUE}
activity$interval<-sprintf("%04d", activity$interval) ## from 0000 to 2355
activity$date<-as.Date(activity$date)
```

Dataframe principal caracteristics are as follows:
```{r showing_df_characteristics, results='markup', echo=TRUE}
summary(activity)
```

It is noteworthy that the following days don't have any observation at all and
contain all the ``r sum(is.na(activity$steps))` NAs` of the dataset:
```{r listing_dates_without_observations, results='markup', echo=TRUE}
## only the dates without any of its 12*24=288 values is considered
print(which(tapply(is.na(activity$step),activity$date, sum)==12*24))
```

The total number of steps per each day is computed and put into the new
dataframe `values.per.day`; this df will be used in the subsequent analysis.
```{r calculating_totalsteps_per_day, results='markup', echo=TRUE}
## 8 days with only NAs are excluded
values.per.day<-aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
## assigns a more proper name to the second variable
names(values.per.day)[2]<-"tot.steps"
```


## What is mean total number of steps taken per day?
The frequencies of the total number of steps per day are distributed as per the following histogram, where is added also a vertical line corresponding to the close values of the median and of the mean of the total number of steps per day.
```{r showing_histogram, results='markup', echo=TRUE}
histogram1<-hist(values.per.day$tot.steps, breaks=12, main="Histogram of total steps per day", xlim=c(0,24000),
     ylim = c(0,20), xlab="Ranges of total number of steps per day", axes=FALSE, col="lightblue")
axis(1,at = seq(0,24000,4000),labels = TRUE,pos = 0)
axis(2,at = seq(0,20,2),labels = TRUE,pos = 0)
mean.tot.steps<-round(mean(values.per.day$tot.steps),1) ## to be used in the plot and in the subsequent question
median.tot.steps<-median(values.per.day$tot.steps) ## to be used in the plot and in the subsequent question
abline(v=mean.tot.steps, col ="red", lwd=2)
text(paste(mean.tot.steps, " = mean value",sep=""),
     x=11500, y=17, adj=c(0,0), cex=0.8, srt=30)
text(paste(median.tot.steps, " = median value",sep=""),
     x=11500, y=16, adj=c(0,0), cex=0.8, srt=30)
```

As already shown in the above histogram the values of the  median = ``r sprintf("%1.0f", median.tot.steps)`` and of the mean =  ``r sprintf("%1.1f", mean.tot.steps)`` of the total number of steps per day are very close


## What is the average daily activity pattern?

The number of steps in each 5'-interval is averaged across all days
(excluding those with NAs) and put into a new dataframe ``daily.pattern``.
```{r analysing_plotting_daily_activity_pattern , results='asis', echo=TRUE}
## creates a variable to host the number of steps in each interval averaged
## across all days - NAs are not considered
daily.pattern<-aggregate(steps ~ interval, data=activity, FUN="mean", na.rm=TRUE)
```

The ``daily.pattern`` is shown in the following line plot, whose x-axis refers
to the intervals from 0000 to 2355, showing only a limited number of them at regular space for sake of clarity

```{r plotting the pattern, results="markup, echo=TRUE"}
plot(daily.pattern$steps, type="l", lwd=2, main="Average number of steps in each 5' interval",
     xlab="", ylab="# of steps", col="blue", axes=FALSE)
axis(1, at=c(1, 6*12, 12*12, 18*12, 24*12),
     labels=daily.pattern$interval[c(1, 6*12, 12*12, 18*12, 24*12)])
axis(2, at=seq(0, 200, by=20), cex.axis=0.8)
## adds 3 horizontal lines at the level of: 5; 100; average number of steps taken in 5' interval
abline(h=20, col="green", lwd=2)
abline(h=100, col="green", lwd=2)
abline(h=mean(daily.pattern$steps), col="red", lwd=2)
text(paste("average = ", round(mean(daily.pattern$steps,1)), sep=""), x=10,
     y=round(mean(daily.pattern$steps,0))+2, adj=c(0,0), cex=0.9, col="red")
```

The 5'-interval with the maximum number of steps on average across all days is the one starting at 0835.
```{r chunk for_identifying_max_5_interval, results='markup', echo=TRUE}
## finds the interval with the maximum number of steps on average
daily.pattern[which.max(daily.pattern$step),]
```

As shown in the plot, on average the activity is very limited (below 5 steps per 5'-interval) from 0000 to 0530 and after 2235
```{r chunk for_analysing_daily_activity_pattern , results='markup', echo=TRUE}
## identifies and shows the set of all the 5'-intervals with less than 5 steps
print(daily.pattern$interval[which(daily.pattern$steps<5)])
```
The activity increments significantly between 0535 and 0605 and even more from 0800 to 0835, when it reaches its maximum value; afterwards it goes under the average value and then oscillates between around 20 and 100 steps per 5'-interval until around 2100, when it starts a descending trend.


## Imputing missing values
The number of rows with missing values (NAs) is quite relevant, both in absolute terms (=``r sum(is.na(activity$step))``) and in relative terms (=``r paste(round(mean(is.na(activity$step)),2)*100,"%", sep="")``).

In order to impute the missing values to the intervals of the 8 dates with NAs,  the average of the corresponding values across the same days of the week are
taken; in this way the typical characteristics of each day of the week are kept.

```{r imputing_missing_values , results='asis'}
## as found at the beginning of the analysis, missing valued are concentrated
## only in some dates, that miss all their value; therefore it is sufficient to
## select the dates and work on them that are stored into a dataframe
uncomplete.dates<-data.frame(date=unique(activity$date[which(is.na(activity$step))]))
## associate the day-of-the-week (var DoW) to each element of "uncomplete.dates"
## df using the function weekdays()
uncomplete.dates<-cbind(uncomplete.dates, DoW=weekdays(uncomplete.dates$date))
## defines a new df, called DoW.avgs (Days-of-Weeks.averages) that in its first
## column contains the 288 5'-intervals and in each of the following 7 columns
## the 288 number of steps averaged across every day of the week, from Monday
## (col 2) to Sunday (col8)
DoW.avgs<-data.frame(interval=unique(activity$interval)) ## initizialized with
                                                         ## the var "interval"
dayofweek=c("Monday", "Tuesday", "Wednesday","Thursday","Friday", "Saturday", "Sunday")
for (i in 1:7){
  ## subsetting the activity dataset per day-of-the-week and eliminating the NAs
  data.ref<-cbind(subset(activity, weekdays(date)==dayofweek[i]& !is.na(activity$steps)))
  ## calculates the average number of steps for each 5'-interval and
  ## stores it into as the (i+1)-th new column of DoW.avgs dataframe
  DoW.avgs<-cbind(DoW.avgs, aggregate(steps ~ interval, data=data.ref, FUN="mean")[2])
}
names(DoW.avgs)=c("interval", paste("avg.step", dayofweek, sep="."))

## creates a copy of the "activity" df and calls it "compl.act"
compl.act<-activity
## completes the missing values of each uncomplete dates of "compl.act"
## with the averaged values of the correspondnt day.of.the.week
for(i in 1:dim(uncomplete.dates[1])[1]){
  compl.act$steps[compl.act$date==uncomplete.dates$date[i]]<-DoW.avgs[,which(dayofweek==uncomplete.dates$DoW[i])+1]
}
```

The total number of steps per each day - including the 8 with reconstructed
values - is computed and put into the new dataframe `compl.values.per.day`.
```{r building_statistics_per_day_on_completed_data, results='markup', echo=TRUE}
compl.values.per.day<-aggregate(steps ~ date, data=compl.act, sum, na.rm=TRUE)
## associates a more proper name to the second variable
names(compl.values.per.day)[2]<-"tot.steps"
```

The frequencies of the total number of steps per day - including the 8 with reconstructed values - are distributed as per the following histogram, where are added also two vertical lines corresponding to the values of the median and of the mean of the total number of steps per day.
```{r showing_histogram_on_completed_data, results='asis', echo=TRUE}
tit="Histogram of total steps per day (reconstructed dataset)"
histogram2<-hist(compl.values.per.day$tot.steps, breaks=12, main=tit, xlim=c(0,24000),
     ylim = c(0,20), xlab="Ranges of total number of steps per day", axes=FALSE, col="lightblue")
axis(1,at = seq(0,24000,4000),labels = TRUE,pos = 0)
axis(2,at = seq(0,20,2),labels = TRUE,pos = 0)
compl.mean.tot.steps<-round(mean(compl.values.per.day$tot.steps),1) ## to be
## used in the plot and in the subsequent question
compl.median.tot.steps<-median(compl.values.per.day$tot.steps) ## to be used
## in the plot and in the subsequent question
abline(v=compl.mean.tot.steps, col ="red", lwd=1)
text(paste(compl.mean.tot.steps, " = mean value",sep=""),
     x=8500, y=17.5, adj=c(0,0), cex=0.8, srt=30)
abline(v=compl.median.tot.steps, col ="green", lwd=1)
text(paste(round(compl.median.tot.steps,0), " = median value",sep=""),
     x=11500, y=17.5, adj=c(0,0), cex=0.8, srt=30)
```

As already shown in the above histogram also in this dataset with missing values reconstructed, the values of the  median = ``r sprintf("%1.0f", compl.median.tot.steps)`` and of the mean =  ``r sprintf("%1.1f", compl.mean.tot.steps)`` of the total number of steps per day are still quite close, but less than those of the dataset with NAs. The new values calculated on the "reconstructed dataset" are greater that the correspondent values calculated on the dataset with NAs.

```{r kable, results = 'asis'}
compare<-data.frame(dataset=c("original","reconstructed", "delta"),
                    median=c(median.tot.steps,compl.median.tot.steps,
                             compl.median.tot.steps-median.tot.steps),
                    mean=c(mean.tot.steps, compl.mean.tot.steps,
                           compl.mean.tot.steps-mean.tot.steps))
kable(compare, digits = 2, align = c("l", rep("r", 2)))
```

This result means that the missing values are more concentrated in those days of the week that, on average, have a greater number of steps. Then imputing the missing values makes the average representation of the activities more adherent to the facts.


## Are there differences in activity patterns between weekdays and weekends?
```{r chunk for showing the differences between wd and we, results='asis'}
Sys.setlocale("LC_TIME", "en_US.UTF-8") ## to set the language to English
```
