---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---


## Loading and preprocessing the data
The data, representing the observations of the number of steps in 5' intervals recorded from
2012-10-01 to 2012-11-30, are stored in "activity.csv" file, present in the working directory;
its data are read into a dataframe called "activity":
```{r chunk for_loading_data, results='markup', echo=TRUE} 
activity<-read.csv("activity.csv", header=TRUE, stringsAsFactor=FALSE)
## change "interval" variable to string making uniform its format
activity$interval<-sprintf("%04d", activity$interval)
activity$date<-as.Date(activity$date) ## transforms the date var from class string to class Date
## sets the language to have week-days names in English
Sys.setlocale("LC_TIME", "en_US.UTF-8")
```

that has the following caracteristics:
```{r chunk for_showing_dataframe_caracteristics, results='markup', echo=TRUE}
print(str(activity))
print(summary(activity))
```

It is noteworthy that the following days don't have any observation at all:
```{r chunk for_listing_dates_without_observations, results='markup', echo=TRUE}
## only the dates without any of its 12*24=288 values is considered
print(which(tapply(is.na(activity$step),activity$date, sum)==12*24))
```

```{r chunk for_building_statistics_per_day, results='markup', echo=TRUE}
## creates dataframe "values.per.day" summing up the total number of steps of each day
## (8 days with only NAs exluded)
values.per.day<-aggregate(steps ~ date, data=activity, sum, na.rm=TRUE)
## associates a more proper name to the second variable
names(values.per.day)[2]<-"tot.steps"
```

## What is mean total number of steps taken per day?
The frequencies of the total number of steps per day ("values.per.day$tot.steps") are distributed as per the following histogram, where is also added a vertical line that corresponds to the two very close values of the median and of the mean of the total number of steps per day.
```{r chunk for_showing_histogram_total_number_steps_per_day, results='asis', echo=TRUE}
histogram1<-hist(values.per.day$tot.steps, breaks=12, main="Histogram of total steps per day", xlim=c(0,24000),
     ylim = c(0,20), xlab="Ranges of total number of steps per day", axes=FALSE, col="lightblue")
histogram1
axis(1,at = seq(0,24000,4000),labels = TRUE,pos = 0)
axis(2,at = seq(0,20,2),labels = TRUE,pos = 0)
mean.tot.steps<-mean(values.per.day$tot.steps) ## to be used in the plot and in the subsequent question
median.tot.steps<-median(values.per.day$tot.steps) ## to be used in the plot and in the subsequent question
abline(v=mean.tot.steps, col ="red", lwd=2)
text(paste(round(mean.tot.steps,1), " = mean value",sep=""),
     x=11500, y=17, adj=c(0,0), cex=0.8, srt=30)
text(paste(round(median.tot.steps,0), " = median value",sep=""),
     x=11500, y=16, adj=c(0,0), cex=0.8, srt=30)
```

As already shown in the above histogram the median and the mean values of the total number of steps per day are very close, in fact their values are respectively:
```{r chunk for printing the values of the mean and the median of total steps per day, results='markup', message=FALSE, rowname=FALSE, echo=TRUE}
cat("          median of the total number of steps taken per day = ", median.tot.steps, "\n",
    "          mean of the total number of steps taken per day = ",round(mean.tot.steps,1), sep="")
```


## What is the average daily activity pattern?

```{r chunk for_plotting_daily_activity_pattern , results='asis', echo=TRUE}
## creates a variable to host the average number of steps in each interval - NAs are not considered
daily.pattern<-aggregate(steps ~ interval, data=activity, FUN=mean, na.rm=TRUE)
## and plot it as a time series
plot(daily.pattern$steps, type="l", lwd=2, main="Average number of steps in each 5' interval",
     xlab="", ylab="# of steps", col="blue", axes=FALSE)
axis(1, at=c(1, 6*12, 12*12, 18*12, 24*12),
     labels=daily.pattern$interval[c(1, 6*12, 12*12, 18*12, 24*12)])
axis(2, at=seq(0, 200, by=20), cex.axis=0.8)
## adds 3 horizontal lines at the level of: 5; 100; average number of steps taken in 5' interval
abline(h=20, col="green", lwd=2)
abline(h=100, col="green", lwd=2)
abline(h=mean(daily.pattern$steps), col="red", lwd=2)
text(paste("average = ", round(mean(daily.pattern$steps,1)), sep=""), x=10,
     y=round(mean(daily.pattern$steps,0))+2, adj=c(0,0), cex=0.9, col="red")
```

The 5'-interval with the maximum number of steps on average across all day is the one starting at 0835.
```{r chunk for_identifying_max_5_interval, results='markup', echo=TRUE}
## finds the interval with the maximum number of steps on average
print(daily.pattern[which.max(daily.pattern$step),]) 
```

As shown in the plot, on average the activity is very limited (below 5 steps per 5'-interval) from 0000 to 0530 and after 2235
```{r chunk for_analysing_daily_activity_pattern , results='markup', echo=TRUE}
## identifies and shows the set of all the 5'-intervals with less than 5 steps
print(daily.pattern$interval[which(daily.pattern$steps<5)])
```
The activity increments significantly between 0535 and 0605 and even more from 0800 to 0835, when it reaches its maximum value; afterwards it goes under the average value and then oscillates between around 20 and 100 steps per 5'-interval until around 2100, when it starts a descending trend.


## Imputing missing values
The number of rows with missing values (NAs) is quite relevant, both in absolute and in relative terms.
```{r chunk for_showing_info_on_dates_without_observations, results='markup', echo=TRUE} 
print(paste("total number of NAs in the dataset: ", sum(is.na(activity$step)), sep=""))
print(paste("total number of NAs in the dataset / dimension of the dataset: ",
            round(mean(is.na(activity$step)),2)*100,"%", sep=""))
```

In order to impute the missing values to the 8 dates with NAs, for each interval the values averaged across the same days of the week are taken

```{r chunk for_imputing_NAs , results='asis'}
## as found at the beginning of the analysis, missing valued are concentrated only in some dates,
## that miss all their value; therefore it is sufficient to select the dates and work on them
uncomplete.dates<-data.frame(date=unique(activity$date[which(is.na(activity$step))]))
## associate the day-of-the-week (var DoW) to each element of "uncomplete.dates" df
uncomplete.dates<-cbind(uncomplete.dates, DoW=weekdays(uncomplete.dates$date))
## defines a new df, called DoW.avgs (Days-of-Weeks.averages) to contain the average number of steps
## in all the 288 5'-intervals per each day of the week
DoW.avgs<-data.frame(interval=unique(activity$interval)) ## initizialized with the var "interval"
dayofweek=c("Monday", "Tuesday", "Wednesday","Thursday","Friday", "Saturday", "Sunday")
for (i in 1:7){
  ## subsetting the activity dataset per day-of-the-week and eliminating the NAs
  data.ref<-cbind(subset(activity, weekdays(date)==dayofweek[i]& !is.na(activity$steps)))
  ## calculates the average number of steps for each 5'-interval and
  ## stores it into a new column of DoW.avgs dataframe
  DoW.avgs<-cbind(DoW.avgs, aggregate(steps ~ interval, data=data.ref, mean)[2])
}
names(DoW.avgs)=c("interval", paste("avg.step", dayofweek, sep="."))

## creates a copy of the "activity" df and calls it "compl.act"
compl.act<-activity
## completes the missing values of each uncomplete dates of "compl.act"
## with the averaged values of the correspondnt day.of.the.week
for(i in 1:dim(uncomplete.dates[1])[1]){
  compl.act$steps[compl.act$date==uncomplete.dates$date[i]]<-DoW.avgs[,which(dayofweek==uncomplete.dates$DoW[i])+1]
}
```

```{r chunk for_building_statistics_per_day_on_completed_data, results='markup', echo=TRUE}
## creates dataframe "compl.values.per.day" summing up the total number of steps of each day
compl.values.per.day<-aggregate(steps ~ date, data=compl.act, sum, na.rm=TRUE)
## associates a more proper name to the second variable
names(compl.values.per.day)[2]<-"tot.steps"
```
The following histogram shows the distribution across the different ranges of total values per day of the completed dataset;
two vertical lines are added, corresponding to the values of the median and of the mean of the total number of steps per day.
```{r chunk for_showing_histogram_on_completed_data, results='asis', echo=TRUE}
tit="Histogram of total steps per day (imputed dataset)"
histogram2<-hist(compl.values.per.day$tot.steps, breaks=12, main=tit, xlim=c(0,24000),
     ylim = c(0,20), xlab="Ranges of total number of steps per day", axes=FALSE, col="lightblue")
axis(1,at = seq(0,24000,4000),labels = TRUE,pos = 0)
axis(2,at = seq(0,20,2),labels = TRUE,pos = 0)
compl.mean.tot.steps<-mean(compl.values.per.day$tot.steps) ## to be used in the plot and in the subsequent question
compl.median.tot.steps<-median(compl.values.per.day$tot.steps) ## to be used in the plot and in the subsequent question
abline(v=compl.mean.tot.steps, col ="red", lwd=1)
text(paste(round(compl.mean.tot.steps,1), " = mean value",sep=""),
     x=8500, y=17.5, adj=c(0,0), cex=0.8, srt=30)
abline(v=compl.median.tot.steps, col ="green", lwd=1)
text(paste(round(compl.median.tot.steps,0), " = median value",sep=""),
     x=11500, y=17.5, adj=c(0,0), cex=0.8, srt=30)
```




## Are there differences in activity patterns between weekdays and weekends?
```{r chunk for showing the differences between wd and we, results='asis'}
Sys.setlocale("LC_TIME", "en_US.UTF-8") ## to set the language to English
```
